Normalisation Example using Spike-in Control as a Standard

### 2017-08-14 examples/Example_001_ER_CTCF.Rmd
Correcting typos.

---
title: "Normalization Example using internal CTCF peaks as a standard"
author: "Andrew Holding"
date: "8/7/2017"
output: pdf_document
---

@@

This is a normalisation using linear regression to CTCF of MCF7 cells chip treameant etc.

@@

## Load convience functions

These functions facilitate the normalisation of data.

```{r brundle}
source('../package/brundle.R')
```

@@

@@

To keep file size down these are provided as a Rdata File rather than as raw counts.

@@

```{r}
#Load Sample Ids from control sample sheet.
jg.sampleIds <- jg.getSampleIds(jg.controlSampleSheet)
```

@@

@@

##Plot showing normalization calculation (Optional)
  jg.plotNormalization(jg.controlCountsTreated,
                     jg.controlCountsUntreated)
```

```{r}
##Get Normalization Coefficient
jg.coefficient<-jg.getNormalizationCoefficient(jg.controlCountsTreated,
                                               jg.controlCountsUntreated)
jg.correctionFactor<-jg.getCorrectionFactor(jg.experimentSampleSheet,
                                            jg.treatedNames,
                                            jg.untreatedNames)

@@

@@

```{r}
#Return values to Diffbind and plot normalised result.
jg.dba <- DiffBind:::pv.resetCounts(dbaExperiment,
                                    jg.experimentPeaksetNormalised)

>>>>

---
title: "Normalisation example using Internal CTCF Peaks as a Standard"
author: "Andrew Holding"
date: "7/8/2017"
output: pdf_document
---

@@

This is a worked example of our method to normalize ER ChIP-seq data using CTCF binding to provide an internal control.

## Preprocessing

To prepare the data for the differential binding analysis, the data had to be aligned and CTCF and ER peaks called. As our CTCF control does not require any special alignment, data was aligned as for a standard ChIP-seq experiment and followed by removal of reads from blacklisted regions. Peak calling was done with macs2 using the script shown below.

```{bash, eval=FALSE}

#Calls peaks for all files in a directory against a single control. Change
# <filename> to the filename of your input sample.

 control=<filename>

 for bam in *bam
 do
   root=`basename $bam .bam`
   macs2 callpeak -t $bam -c $control -f BAM -n $root -g hs
 done

As macs2 cannot differentiate between a peak arising from CTCF binding or ER binding, we used samples  prepared with only CTCF antibody to establish a CTCF binding consensus: one treated and one untreated. These peaks were then merged as follows.

```{bash, eval=FALSE}

cat  SLX-14438.D701_D504.HKN27BBXX.s_8.r_1.fq.gz.bam_peaks.xls \
      SLX-14438.D702_D503.HKN27BBXX.s_8.r_1.fq.gz.bam_peaks.xls > CTCF_merged.bed

grep -v \# CTCF_merged.bed  | grep -v start > CTCF_merged_filtered.bed
sort -k1,1 -k2,2n CTCF_merged_filtered.bed > CTCF_merged_sorted_filtered.bed

bedtools merge -i CTCF_merged_sorted_filtered.bed  > CTCF_union.bed

We then used this CTCF consensus to establish specific CTCF or ER binding in the experimental samples, either by taking the intersection of the CTCF consensus and the peak file for a particluar sample to provide a list of CTCF binding sites in the sample, or by subtratracting the consensus CTCF peaks from the macs2 output to provide a list of ER binding sites for that sample.

```{bash, eval=FALSE}

mkdir CTCF
for f in *.narrowPeak
do
 	echo $f
 	bedtools intersect -a $f -b CTCF_union.bed  > CTCF/$f 
done

mkdir ER
for f in *.narrowPeak
do
 	echo $f
 	bedtools subtract -b CTCF_union.bed -a $f -A > ER/$f 
done

The peak list files were then used as the basis for two DiffBind sample sheets as described in the DiffBind vignette. One uses the CTCF peak list generated from the above scripts; this is the control sample sheet. The second uses the ER peaks; this is the experimental sample sheet. These two sample sheets are then provided as the input in the analysis below.

@@

## Load convenience functions

Brundle provies a series of functions we developed for this pipeline that are used to normalise ChIP-seq data. The usage of these functions are explained in the example that follows.

```{r brundle,results='hide', message=FALSE, warning=FALSE}
source('../package/brundle.R')
```

@@

The intial settings for the analysis are minimal. The variable jg.controlMinOverlap is set to 5 as this is 1 less than the total sample count and we want a strong consensus of CTCF peaks to normalise against. The control and experimental sample sheets are as described above. The conditions reflect the treated and untreated conditions we have used when writing our sample sheet.

@@

Typically we would load the raw data using DiffBind as we are planning to return our normalised data as a DiffBind object.

In this example, as the BAM files are greater in size than the GitHub file size limit, we have included the data as the DiffBind object that these functions would output and instead of loading from the raw data directly. If the code does not find the R data object, it will try to regenerate it from the raw data.

@@

DiffBind allows us to extract the counts in each peak. We use this function to obtain the data in a form that is easier to normalise.

@@

To normalise, we need to separate samples out by condition. This is done using the following function. The peakset contains the reads in peaks; the sample sheet provides the metadata needed and the jg.treatedCondition/jg.untreatedCondition specifies which condition we want to get the counts for.

@@

The normalisation factor is generated by a linear regression from the mean number of counts for each peak before and after treatment. The gradient of the linear fit should be equal to 1 given the assumption that, before and after treatment, the majority of the control peaks do not change.

This plot is not needed to calculate the normalisation factor but it does provide a visual interpretation of the process.

```{r, message=FALSE}

# Plot showing normalisation calculation (Optional)

jg.plotNormalization(jg.controlCountsTreated,
                     jg.controlCountsUntreated)
```

There are two parts to the normalisation coefficent we use in the final correction. The first is the the coefficent to make the levels of CTCF binding before and after treament equal. The second is because DiffBind normalises against the total read count for each sample, we therefore generated a correction factor to control for this.

```{r}
# Get Normalisation Coefficient

jg.coefficient<-jg.getNormalizationCoefficient(jg.controlCountsTreated,
                                               jg.controlCountsUntreated)
jg.correctionFactor<-jg.getCorrectionFactor(jg.experimentSampleSheet,
                                            jg.treatedNames,
                                            jg.untreatedNames)

@@

Once we have these two peices of information, we apply both the normalisation coeffficient and the correction factor to our data.

@@

Finally we return the correct values into the DiffBind object. This requires a modified version DiffBind as included in the original manuscript. It is expected that this will be supported natively by future versions of DiffBind.

```{r}
#Return values to DiffBind and plot normalised result.
jg.dba <- DiffBind:::pv.resetCounts(dbaExperiment,
                                    jg.experimentPeaksetNormalised)

### 2017-08-14 examples/Example_002_ER_CTCF_DeSeq.Rmd
Edited examples
Correcting typos.

---
title: "Normalization Example using internal CTCF peaks as a standard (no-linear fit)"
author: "Andrew Holding"
date: "8/7/2017"
output: pdf_document
---

@@

This is a worked example of our method alternative to normalize ER ChIP-seq data using CTCF binding to provide an internal control. Unlike our previous example this normalisation is using only DESeq Size Factors. This method was found to be less effective but was used for comparision in the supporting publication. This markdown builds on Example 001 and therefore that should be read first. Example 001 also explains the pre-processing steps.

@@

## Load convience functions

@@

In this example Diffbind was simply used as a convient way to extract peak counts and provided consitancy for comparision between methods.

To keep file size down counts are provided as a Rdata file rather than as raw data.

@@

The actual DESeq Pipeline starts simply needs a matrix so we covert from Diffbind and save as csv to provide a starting point.

@@

filename<-"csv/experimentalPeakset.csv"
if(!file.exists(filename)){
  ## Extract Peak set from DiffBind
  jg.experimentPeakset <- jg.dbaGetPeakset(dbaExperiment)
  #Convert Peakset to DeSeq Workflow
  jg.experimentPeaksetDeSeq<-jg.convertPeakset(jg.experimentPeakset)
  #Save to CSV

@@

Here we use the CSV files the example code above generated; however, it is possible to generate the input data through a wide variety of methods and we have provided only one.

@@

To normalize the data we use the DESeq2 function to estimate the size factors. This is a very simple method, and it provides some degree of normalization. We also use the original sample sheet to get the different conditions.

@@

#Establish size factors directly from Control data
jg.controlSizeFactors = estimateSizeFactorsForMatrix(jg.controlPeaksetDeSeq)

@@

The Brundle package also contains some convience functions for plotting the output of DESeq2 inplace of DiffBind. These generate the following figures.

>>>>

---
title: "Normalisation Example using Internal CTCF Peaks as a Standard (No-linear Fit)"
author: "Andrew Holding"
date: "7/8/2017"
output: pdf_document
---

@@

This is a worked example of our alternative method to normalise ER ChIP-seq data using CTCF binding to provide an internal control. Unlike our previous example, this normalisation is using only DESeq Size Factors. This method was found to be less effective but was used for comparision in the supporting publication. This markdown builds on Example 001 and therefore that should be read first. Example 001 also explains the pre-processing steps.

@@

## Load convenience functions

@@

In this example DiffBind was simply used as a convenient way to extract peak counts and provided consisteancy for comparision between methods.

To keep file sizes down, counts are provided as a Rdata file rather than as raw data.

@@

The actual DESeq Pipeline starts simply needs a matrix so we convert from DiffBind and save as a CSV file to provide a starting point.

@@

filename<-"csv/experimentalPeakset.csv"
if(!file.exists(filename)){
  ## Extract Peak set from DiffBind
  jg.experimentPeakset <- jg.dbaGetPeakset(dbaExperiment)
  #Convert Peakset to DESeq Workflow
  jg.experimentPeaksetDeSeq<-jg.convertPeakset(jg.experimentPeakset)
  #Save to CSV

@@

Here we use the CSV files that the example code above generated; however, it is possible to generate the input data through a wide variety of methods and we have provided only one.

@@

To normalize the data, we use the DESeq2 function to estimate the size factors. This is a very simple method, and it provides some degree of normalization. We also use the original sample sheet to get the different conditions.

@@

#Establish size factors directly from control data
jg.controlSizeFactors = estimateSizeFactorsForMatrix(jg.controlPeaksetDeSeq)

@@

The Brundle package also contains some convenience functions for plotting the output of DESeq2 in place of DiffBind. These generate the following figures.

### 2017-08-14 examples/Example_003_ER_H2av.Rmd
Correcting typos.

---
title: "Normalization Example using spike-in control as a standard"
author: "Andrew Holding"
date: "8/7/2017"
output: pdf_document
---

@@

This is a normalisation using H2av binding within an external spike-in of droshilia chromatin to provide a control for the analysis of ER binding in MCF7 cells after treatment with Fulvestrant.

This example builds on the previous two and therefore these should be read first as many details of key functions have not been repeated.

@@

Unlike Example 001 & 002 the use of a xenogenic spike in requires more pre-processing. The scripts for this process are provided below. Genomes were downloaded from http://support.illumina.com/sequencing/sequencing_software/igenome.html.

The first step of our preprossing pipeline requires the generation of a combined Human/Drosophila genome. We used the following script to generated this. The human and drosophila genomes downloaded from the link provided above are contained in a subdirectory "genomes" relative to where the script was run. These example scripts are not evalutated in this markdown due to the time  it takes to process the data.

@@

To allow use to use the merged genome we first indexed it using bowtie2. For our system we specified 32 threads, this will depend on the number of processors you have available.

@@

Once indexed, we could the aligned to the merge genome. This script assumes that the FastQ files are in a directory ./SLX-8047_dhms from where the scripts are run.

```{bash, eval=FALSE}
#Align, Sort and Build indeces

@@

After aligning the data can calling peaks we can then split the data back into Drosphila and Human.

@@

With the peak files and the seperate alignments extract we are can then write samplesheets. For the normalization we need one for the drosphila aligned reads from each sample and the respective H2av peak file, and one for human reads from each sample and their repsective ER peak files. This forms the bais of the following analysis.

## Load convience functions

@@

As before to keep file size down data are provided as a RdatafFile rather than as raw BAM files.

@@

##Plot showing normalization calculation (Optional)
  jg.plotNormalization(jg.controlCountsTreated,
                     jg.controlCountsUntreated)
```

```{r}
##Get Normalization Coefficient
jg.coefficient<-jg.getNormalizationCoefficient(jg.controlCountsTreated,
                                               jg.controlCountsUntreated)
jg.correctionFactor<-jg.getCorrectionFactor(jg.experimentSampleSheet,
                                            jg.treatedNames,
                                            jg.untreatedNames)

@@

```{r}
#Return values to Diffbind and plot normalised result.
jg.dba <- DiffBind:::pv.resetCounts(dbaExperiment,
                                    jg.experimentPeaksetNormalised)

>>>>

---
title: "Normalisation Example using Spike-in Control as a Standard"
author: "Andrew Holding"
date: "7/8/2017"
output: pdf_document
---

@@

This is a normalisation using H2av binding within an external spike-in of drosophilia chromatin to provide a control for the analysis of ER binding in MCF7 cells after treatment with Fulvestrant.

This example builds on the previous two examples and therefore they should be read first as many details of key functions have not been repeated here.

@@

Unlike Examples 001 and 002 the use of a xenogenic spike-in requires more preprocessing. The scripts for this process are provided below. Genomes were downloaded from http://support.illumina.com/sequencing/sequencing_software/igenome.html.

The first step of our preprossing pipeline requires the generation of a combined human/drosophila genome. We used the following script to generate this. The human and drosophila genomes downloaded from the link provided above are contained in a subdirectory "genomes" relative to where the script was run. These example scripts are not evalutated in this markdown due to the time it takes to process the data.

@@

To allow us to use the merged genome we first indexed it using bowtie2. For our system, we specified 32 threads; the number of threads will depend on the number of processors you have available.

@@

Once indexed, we could then align to the merged genome. This script assumes that the FastQ files are in a directory ./SLX-8047_dhms from where the scripts are run.

```{bash, eval=FALSE}
#Align, Sort and Build indexes

@@

After aligning the data and calling peaks, we can then split the data back into human and drosophila.

@@

With the peak files and the separate alignments extracted, we can then write sample sheets. For the normalization we need one for the drosophila aligned reads from each sample and the respective H2av peak file, and one for human reads from each sample and their respsective ER peak files. This forms the bais of the following analysis.

## Load convenience functions

@@

As before, to keep file size down, data are provided in Rdata format rather than as raw BAM files.

@@

##Plot showing normalisation calculation (Optional)
  jg.plotNormalization(jg.controlCountsTreated,
                     jg.controlCountsUntreated)
```

```{r}
##Get Normalisation Coefficient
jg.coefficient<-jg.getNormalizationCoefficient(jg.controlCountsTreated,
                                               jg.controlCountsUntreated)
jg.correctionFactor<-jg.getCorrectionFactor(jg.experimentSampleSheet,
                                            jg.treatedNames,
                                            jg.untreatedNames)

@@

```{r}
#Return values to DiffBind and plot normalised result.
jg.dba <- DiffBind:::pv.resetCounts(dbaExperiment,
                                    jg.experimentPeaksetNormalised)
